The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-10-28 categorical-zero-one.tex

	If we do not yet assume that $X_{J \,\sqcup\, \{\ast\}}$ exists, then we can exhibit $X_J \otimes X_\ast$ as the relevant cofiltered limit as follows. The object $X_J \otimes X_\ast$ can be equipped finite marginalization morphisms with respect to $F \subseteq J \sqcup \{\ast\}$ given by
	\[
		\pi_{F \setminus \{\ast\}} \otimes \id \: : \: X_J \otimes X_\ast \longrightarrow X_{F \setminus \{\ast\}} \otimes X_\ast
	\]
	for $\ast \in F$, and just
	\[
		\pi_F \otimes \discard{X_\ast} : X_J \otimes X_\ast \longrightarrow X_F 
	\]
	in case $\ast \not \in F$. It is straightforward to check that these morphisms equip $X_J \otimes X_\ast$ with the right universal property of the infinite tensor product $X_{J \sqcup \{\ast\}}$ if only if the functor $- \otimes X_\ast$ preserves the defining cofiltered limit of the infinite tensor product $X_J$. It is also easy to see that this new cofiltered limit is again preserved by every functor $- \otimes Y$, based on the assumption that, based on the assumption that the defining cofiltered limit of the original infinite tensor product is preserved by $- \otimes (X_\ast \otimes Y)$ as well.

@@

\begin{example}
	\label{infprods_stoch}
	We now consider infinite tensor products in $\Stoch$, where we would like infinite tensor products to be given by the corresponding infinite products of measurable spaces in the usual sense. Since the Kolmogorov extension theorem does not hold for general (even merely countable) products of measurable spaces~\cite{AJ}, this is not the case without further additional assumptions on the measurable spaces involved. Thus we do not know whether $\Stoch$ has infinite tensor products, although we suspect that it does not; but even if it does, they are in general not the ones that one would like to have for the purposes of probability theory.

@@

\todo[inline]{Add an example where $\cC_{\rm det}$ has infinite products, but $\cC$ lacks infinite tensor products? Like sets and multivalued functions?}

\todo[inline]{T: sounds good to me, can you put it in?}

@@

\begin{lemma}[The determinism lemma]
	\label{determinismlemma}
	Suppose that we have $p: A \to X$ and deterministic $s : X \to T$. If the joint
	\begin{equation}
		\label{determinismlemma_joint}
		\tikzfig{determinismlemma_joint}	
	\end{equation}
	displays the conditional independence $X \perp T \mid\mid A$, then the composite $sp: A \to T$ is deterministic.
\end{lemma}
This result can be thought of as a partial converse to the condition which characterizes positivity of a Markov category~\cite[Definition~11.22]{markov_cats}. Here and in the following, our notation ``$s$'' indicates that we think of this morphism as a statistic~\cite[Definition~14.2]{markov_cats}.
\begin{proof}
	Since the marginals of \Cref{determinismlemma_joint} are exactly $p$ and $sp$, the conditional independence assumption is
	\[
		\tikzfig{determinismlemma1}
	\]
	Using this equation together with the assumption that $s$ is deterministic, we then compute
	\[
		\tikzfig{determinismlemma2}
	\]
	which is what was to be shown.
\end{proof}

>>>>

	If we do not yet assume that $X_{J \,\sqcup\, \{\ast\}}$ exists, then we can exhibit $X_J \otimes X_\ast$ as the relevant cofiltered limit as follows. The object $X_J \otimes X_\ast$ can be equipped finite marginalization morphisms with respect to $F \subseteq J \sqcup \{\ast\}$ given by
	\[
		\pi_{F \setminus \{\ast\}} \otimes \id \: : \: X_J \otimes X_\ast \longrightarrow X_{F \setminus \{\ast\}} \otimes X_\ast
	\]
	for $\ast \in F$, and just
	\[
		\pi_F \otimes \discard{X_\ast} : X_J \otimes X_\ast \longrightarrow X_F 
	\]
	in case $\ast \not \in F$. It is straightforward to check that these morphisms equip $X_J \otimes X_\ast$ with the right universal property of the infinite tensor product $X_{J \sqcup \{\ast\}}$ if and only if the functor $- \otimes X_\ast$ preserves the defining cofiltered limit of the infinite tensor product $X_J$. It is also easy to see that this new cofiltered limit is again preserved by every functor $- \otimes Y$, based on the assumption that the defining cofiltered limit of the original infinite tensor product is preserved by $- \otimes (X_\ast \otimes Y)$ as well.

@@

\begin{example}

	\label{infprods_stoch}
	We now consider infinite tensor products in $\Stoch$, where we would like infinite tensor products to be given by the corresponding infinite products of measurable spaces in the usual sense. Since the Kolmogorov extension theorem does not hold for general (even merely countable) products of measurable spaces~\cite{AJ}, this is not the case without further additional assumptions on the measurable spaces involved. Thus we do not know whether $\Stoch$ has infinite tensor products, although we suspect that it does not; but even if it does, they are in general not the ones that one would like to have for the purposes of probability theory.

@@

\begin{example}
    \label{setmulti}
    Let $\SetMulti$ be the category with objects ordinary sets, maps $f: X \to Y$ given by functions (i.e maps in the ordinary category of sets) $f: X \to \cP(Y)\setminus \emptyset$, and composition given by
    \[f\circ g(x) = \bigcup_{y \in g(x)} f(y)\]
    for $g: X \to Y, f: Y \to Z$.
    We think of the morphisms in $\SetMulti$ as \emph{multivalued functions}.
    See also \cite[Example~2.6]{markov_cats}.
    The Cartesian product of sets gives it a symmetric monoidal structure, if we set \[(f \tensor g)(x_1,x_2) = f(x_1) \times g(x_2) \subset Y_1 \times Y_2,\] for $f:X_1 \to Y_1, g: X_2 \to Y_2$.
    This is a semicartesian monoidal structure, and we can give it a Markov category structure by setting
    $\cop{X}(x) = \{(x,x)\} \subset X \times X$.
    Then the deterministic morphisms are precisely those sending each point into a singleton set, and there is an obvious isomorphism of categories
    $\SetMulti_\detc \cong \sf{Set}$.
    Of course, $\sf{Set}$ actually has all infinite products, but they do not form Kolmogorov products, or even infinite tensor products in $\sf{SetMulti}$.
    To see this, consider the family $X_n = \{0,1\}$ for $n \in \bN$.
    A map $* \to \prod_{n\in \bN} \{0,1\} = X_\bN$ is just a subset of this infinite product.
    Given such a subset $U$, the finite marginals $* \to X_F$ for $F \subset \bN$ are the images of $U$ under the projection map.
    But these images do not determine $U$. For instance, consider the subset $U$ of those sequences containing at least one $1$.
    The image in $X_F$ is all of $X_F$ (since any finite sequence of $0$s and $1$s can be extended to one containing a $1$).
    But $U$ is not all of $X_\bN$.

\end{example}

@@

\begin{lemma}[The determinism lemma]
	\label{determinismlemma}
	Suppose that we have $p: A \to X$ and deterministic $s : X \to T$. If the joint
	\begin{equation}
		\label{determinismlemma_joint}
		\tikzfig{determinismlemma_joint}	
	\end{equation}
	displays the conditional independence $X \perp T \mid\mid A$, then the composite $sp: A \to T$ is deterministic.
\end{lemma}\cite
This result can be thought of as a partial converse to the condition which characterizes positivity of a Markov category~\cite[Definition~11.22]{markov_cats}. Here and in the following, our notation ``$s$'' indicates that we think of this morphism as a statistic~\cite[Definition~14.2]{markov_cats}.
\begin{proof}
	Since the marginals of \Cref{determinismlemma_joint} are exactly $p$ and $sp$, the conditional independence assumption is
	\[
		\tikzfig{determinismlemma1}
	\]
	Using this equation together with the assumption that $s$ is deterministic, we then compute
	\[
		\tikzfig{determinismlemma2}
	\]
	which is what was to be shown.
\end{proof}
